Classless Java

### 2015-11-21 paper/sections/Overview.tex
More polishing on Section 2

\subsection{Simple Traits with Default Methods}

@@

The animal system modelled so far, is a simplified version of the
system presented in Figure~\ref{}. \bruno{emphasize that, so far the
  need for \mixin is not justified?} Lets now move on to modelling 
the complete animal system. In order to do this, animals will include 
a notion of points, representing their location in space.

@@

However we are not required to use this cumbersome
syntax\footnote{Available in Java from version ...} for every object
allocation.  As programmers do, for ease or reuse, we can encapsulate
boring long repetitive code in a method. A static factory method in
this case is appropriate:
\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int x, int y){return new Point2D(){
    public int x(){return x;} public int y(){return y;}};}
  }
\end{lstlisting}

\paragraph{\Q@Point2D@ with Object Interfaces}
This obvious constructor code can be automatically generated by our
\mixin annotation.  By annotating the interface \Q@Point2D@, the
annotation will generate a variation of the shown static method
\texttt{of}, mimicking the functionality of a simple minded
constructor: By looking at the methods that need implementations it
first detects what are the fields, then generates an \Q@of@ method with
one argument for each of them.That is, we can just write

@@

More precisely, a field or factory parameter is generated for every
no-args method requiring implementation whose name does not have
special meaning\footnote{Formal definition of \mixin behaviour is
  provided later.}.  An example of code using \Q@Point2D@ is
\Q@Point2D.of(42,myPoint.y())@ where we return a new point, using
\Q@42@ as x-coordinate, and taking all the other informations (only
\Q@y@ in this case) from another point.

\paragraph{\texttt{with} methods in Object Interfaces}
This pattern is very common
when programming with immutable data-structures; it is so common that
we decided to support it in our code generation as \Q@with-@ methods,
that is:
\begin{lstlisting}
@Mixin interface Point2D{ int x();    int y();
  Point2D withX(int val);    Point2D withY(int val);}
\end{lstlisting}
is expanded by our annotation into\footnote{
Note how we actually generate a real field \Q@int x=_x;@.
This provide a more uniform translation that can work also for mutable datastructures, where setters are required.
}

\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int _x, int _y){return new Point2D(){
    int x=_x; int y=_y;
    public int x(){return x;}   public int y(){return y;}
    Point2D withX(int val){ return Point2D.of(val,this.y());}    
    Point2D withY(int val){ return Point2D.of(this.x(),val);}  
    };}
  Point2D withX(int val);    Point2D withY(int val);}
\end{lstlisting}
If the programmer would like a different implementation, they can just provide it using \Q@default@; that is:
\begin{lstlisting}
@Mixin interface Point2D{ int x(); int y();
  default Point2D withX(int val){ ... };    default Point2D withY(int val){ ... }}
\end{lstlisting}
is expanded into
\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int _x, int _y){return new Point2D(){
    int x=_x;    int y=_y;
    public int x(){return x;}    public int y(){return y;}
    };}
  default Point2D withX(int val){ ... }    default Point2D withY(int val){ ... }}
\end{lstlisting}

That is, we can provide implementation for methods in an interface, 
and our annotation generate code only if the method \emph{needs} implementation. In this way is trivial for the programmer to personalize the behaviour if they have any special need.

@@

\subsection{\Q@Animal@ and \Q@Horse@: simple mutable data with fields}
\Q@Points2D@ are mathematical entities, thus we chosen to use immutable data structure to model them; however animals are real world entities, and when an animal moves, it is the same animal that have now a different location. We model this with mutable state.
%Now we proceed to define \texttt{Animal} with \texttt{point} ``member
%field''.
%%Not again, before we used only getters!
% Again, we model this member field with getter and setter methods:
\lstinputlisting[linerange=64-67]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=ANIMAL
\marco{TODO: change point into ``location''}
Here we declare abstract getter and setter for the mutable ``field'' \Q@location@.
As you see, we do not apply the \mixin annotation. This is morally equivalent to an abstract class in full Java; that is we do not provide a convenient way to instantiate it.

\lstinputlisting[linerange=71-76]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=HORSE

For \texttt{Horse}, concrete implementation of \texttt{run()} method
need to be defined in a default method, where 
we also show the convenience of \emph{with} methods.
% by method
%\texttt{run()}: method \texttt{withX} returns a new point object with field
%\texttt{x} updated by the argument to \texttt{withX}. Without these
%\texttt{with} methods, operations like \texttt{run()} would be much harder to define.

@@

\lstinputlisting[linerange=80-88]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=BIRD

@@

\lstinputlisting[linerange=96-97]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=PEGASUS

@@

\lstinputlisting[linerange=71-76]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=HORSE

\lstinputlisting[linerange=80-88]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=BIRD

\lstinputlisting[linerange=96-97]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=PEGASUS

>>>>

\subsection{Simple Multiple Inheritance with Default Methods}

@@

The animal system modelled so far, is a simplified version of the
system presented in Figure~\ref{}, and still does not appear to 
justify the \mixin annotation. Lets now move on to modelling 
the complete animal system. In order to do this, animals will include 
a notion of points, representing their location in space. As we shall
see, modelling stateful components using plain Java 8 quickly becomes 
cumbersome. The \mixin annotation comes to rescue here and 
allows avoiding significant amounts of boilerplate code.

@@

However programmers are not required to use this cumbersome syntax for
every object allocation. As programmers do, for ease or reuse, the
boring long repetitive code can be encapsulate in a method. A
generalization of the \texttt{of} static factory
method is appropriate in this case:
\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int x, int y){return new Point2D(){
    public int x(){return x;} public int y(){return y;}};}
}
\end{lstlisting}

\vspace{-5pt}
\paragraph{\Q@Point2D@ with Object Interfaces}
This obvious constructor code can be automatically generated by our
\mixin annotation.  By annotating the interface \Q@Point2D@, the
annotation will generate a variation of the shown static method
\texttt{of}, mimicking the functionality of a simple minded
constructor: By looking at the methods that need implementations it
first detects what are the fields, then generates an \Q@of@ method with
one argument for each of them. That is, we can just write

@@

\noindent More precisely, a field or factory parameter is generated for every
no-args method requiring implementation (except for methods whose name have
special meaning).  An example of code using \Q@Point2D@ is:

\begin{lstlisting}
Point2D p = Point2D.of(42,myPoint.y());
\end{lstlisting}

\noindent where we return a new point, using \Q@42@ as x-coordinate,
and taking all the other informations (only \Q@y@ in this case) from
another point.

\paragraph{\texttt{with} methods in Object Interfaces}
The pattern of creating a new piece of data by reusing most
information from an old piece of data is very common
when programming with immutable data-structures; it is so common that
we decided to support it in our code generation as \Q@with@ methods,
that is:
\begin{lstlisting}
@Mixin interface Point2D { 
  int x();    int y(); // getters
  Point2D withX(int val);    Point2D withY(int val); // with methods
}
\end{lstlisting}

\noindent Using \texttt{with} methods, the point \texttt{p} could have
been created instead using:

\begin{lstlisting}
Point2D p = myPoint.withX(42);
\end{lstlisting}

\noindent If there is a large number of fields, \texttt{with} methods
will save programmers from writting large amounts of tedious code that
simply copies the values of fields. 
\begin{comment}
is expanded by our annotation into\footnote{
Note how we actually generate a real field \Q@int x=_x;@.
This provide a more uniform translation that can work also for mutable datastructures, where setters are required.
}

\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int _x, int _y){return new Point2D(){
    int x=_x; int y=_y;
    public int x(){return x;}   public int y(){return y;}
    Point2D withX(int val){ return Point2D.of(val,this.y());}    
    Point2D withY(int val){ return Point2D.of(this.x(),val);}  
    };}
  Point2D withX(int val);    Point2D withY(int val);}
\end{lstlisting}
\end{comment}
Moreover, if the programmer wants a different implementation, they
just provide it using a \Q@default@ method. For example:
\begin{lstlisting}
@Mixin interface Point2D{ 
  int x(); int y();
  default Point2D withX(int val){ ... };    default Point2D withY(int val){ ... }}
\end{lstlisting}

\noindent can be used by programmers to define their own behaviour for 
the \texttt{with} methods.

\begin{comment}
is expanded into
\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int _x, int _y){return new Point2D(){
    int x=_x;    int y=_y;
    public int x(){return x;}    public int y(){return y;}
    };}
  default Point2D withX(int val){ ... }    default Point2D withY(int val){ ... }}
\end{lstlisting}

That is, we can provide implementation for methods in an interface, 
and our annotation generate code only if the method \emph{needs} implementation. In this way is trivial for the programmer to personalize the behaviour if they have any special need.
\end{comment}

@@

\paragraph{\Q@Animal@ and \Q@Horse@: simple mutable data with fields}
Two dimensional points are mathematical entities, thus we chosen to
use immutable data structure to model them. However animals are real
world entities, and when an animal moves, it is the same animal that
now has a different location. We model this with mutable state.

%Now we proceed to define \texttt{Animal} with \texttt{point} ``member
%field''.
%%Not again, before we used only getters!
% Again, we model this member field with getter and setter methods:
\lstinputlisting[linerange=64-67]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=ANIMAL

\noindent Here we declare abstract getter and setter for the mutable ``field'' \Q@location@.
The \mixin annotation is not used. This is morally
equivalent to an abstract class in full Java; and there is no
convenient way to instantiate it.
For \texttt{Horse}, the \mixin annotation is used and a concrete implementation of \texttt{run()} method
is defined using a default method, which also further illustrates the convenience of \emph{with} methods.

\lstinputlisting[linerange=71-74]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=HORSE

% by method
%\texttt{run()}: method \texttt{withX} returns a new point object with field
%\texttt{x} updated by the argument to \texttt{withX}. Without these
%\texttt{with} methods, operations like \texttt{run()} would be much harder to define.

@@

\lstinputlisting[linerange=78-86]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=BIRD

@@

\lstinputlisting[linerange=94-95]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=PEGASUS

@@

\lstinputlisting[linerange=71-74]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=HORSE

\lstinputlisting[linerange=78-86]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=BIRD

\lstinputlisting[linerange=94-95]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=PEGASUS
